A Gaussian process model of human electrocorticographic data

### 2020-01-28 paper/main.tex

We first applied fourth order Butterworth notch filter to remove 60 Hz
($\pm$ .5 Hz) line noise from every recording (from every electrode).
Next, we downsampled the recordings (regardless of the original
samplerate) to 250 Hz.  (This downsampling step served to both
normalize for differences in sampling rates across patients and to
ease the computational burden of our subsequent analyses.)  We then
excluded any electrodes that showed putative epileptiform activity.
Specifically, we excluded from further analysis any electrode that
exhibited an maximum kurtosis of 10 or greater across all of that
patient's recording sessions.  We also excluded any patients with
fewer than 2 electrodes that passed this criteria, as the SuperEEG
algorithm requires measuring correlations between 2 or more electrodes
from each patient.  For Dataset 1, this yielded clean recordings from
4168 electrodes implanted throughout the brains of 67 patients
(Fig.~\ref{fig:methods}A); for Dataset 2, this yielded clean
recordings from 3159 electrodes from 24 patients.  Each individual
patient contributes electrodes from a limited set of brain locations,
which we localized in a common space~\citep[MNI152;][]{GrabEtal06}; an
example Dataset 1 patient's 54 electrodes that passed the above
kurtosis threshold test are highlighted in black and red.

@@

A second mechanism whereby a given region $x$ can contribute to the
recording at $\eta$ is through (direct or indirect) anatomical
connections between structures near $x$ and $\eta$.  Although
anatomical and function correlations can differ
markedly~\citep[e.g.,][]{AdacEtal11, HoneEtal09, GoniEtal14}, we use temporal
correlations in the data to estimate these anatomical
connections~\citep{BeckEtal18}.  Let $\bar{R}$ be the set of locations
at which we wish to estimate local field potentials, and let
$R_{s} \subseteq \bar{R}$ be set of locations at which we observe local
field potentials from patient $s$ (excluding the electrodes that did
not pass the kurtosis test described above). In the analyses below we
define $\bar{R} = \cup_{s=1}^S R_s$.  We can calculate the expected
inter-electrode correlation matrix for patient $s$, where
$C_{s,k}(i,j)$ is the correlation between the time series of voltages
for electrodes $i$ and $j$ from subject $s$ during session $k$, using:

@@

We designed our approach to be agnostic to electrode impedances, as
electrodes that do not exist do not have impedances.  Therefore our
algorithm recovers voltages in standard deviation ($z$-scored) units
rather than attempting to recover absolute voltages. (This property
reflects the fact that $\hat{K}_{\beta_s, \alpha_s}$ and
$\hat{K}_{\alpha_s, \alpha_s}$ are correlation matrices rather than
covariance matrices.)  Also, we note that
Equation~\ref{eqn:reconstruction} requires computing a $T$ by $T$
matrix, which can become computationally intractable when $T$ is very
large (e.g., for an example patient in Dataset 1, $T =
12786750$). However, because Equation~\ref{eqn:reconstruction} is time
invariant, we may compute $Y_{s,k,\beta_s}$ in a piecewise
manner by filling in $Y_{s,k,\beta_s}$ one row at a time (using the
corresponding samples from $Y_{s, k, \alpha_s}$).

@@

Although both datasets we examined provide good full-brain coverage
(when considering data from every patient; e.g.\
Fig.~\ref{fig:corrmap}C, D), electrodes are not placed uniformly
throughout the brain.  For example, electrodes are more likely to be
implanted in regions like the medial temporal lobe (MTL), and are
rarely implanted in occipital cortex (Fig.~\ref{fig:density}A, B).
Separately for each dataset, for each voxel in the 4~mm$^3$ voxel
MNI152 brain, we computed the proportion of electrodes in the dataset
that were contained within a 20 MNI unit radius sphere centered on
that voxel.  We defined the \textit{density} at that location as this
proportion.  Across Datasets 1 and 2, the electrode placement
densities were similar (correlation by voxel:
$r = 0.6, p < 10^{-10}$).  We wondered whether regions with good
covererage might be associated with better reconstruction accuracy
(e.g. Fig.~\ref{fig:corrmap}C, D indicate that many electrodes in the
MTL have relatively high reconstruction accuracy, and occipital
electrodes tend to have relatively low reconstruction accuracy).  To
test whether this held more generally across the entire brain, for
each dataset we computed the electrode placement density for each
electrode from each patient (using the proportion of \textit{other}
patients' electrodes within 20 MNI units of the given electrode).  We
then correlated these density values with the across-patient
reconstruction accuracies for each electrode.  We found a negative
correlation between reconstruction accuracy and density for Dataset 1,
but was unreliable for Dataset 2 (Dataset 1: $r = -0.21, p = 0.05$; Dataset 2:
$r = -0.30, p = 0.15$).  This indicates that the reconstruction
accuracies we observed are not driven by sampling density, but
rather may also reflect higher order properties of neural dynamics
such as functional correlations between distant
voxels~\citep{BetzEtal17b}.

@@

To gain insights into whether particular electrode locations might be
especially informative, we first computed the average reconstruction
accuracy across all of each patient's electrodes (using the
across-patients cross validation test; black histograms in
Fig.~\ref{fig:corrmap}A and B).  We labeled each patient's electrodes
in each dataset with the average reconstruction accuracy for that
patient.  In other words, we assigned every electrode from each given
patient the same value, reflecting how well the activity patterns at
those electrodes were reconstructed on average.  Next, for each voxel
in the 4~mm$^3$ MNI brain, we computed the average value across any
electrode (from any patient) that came within 20 MNI units of that
voxel's center.  Effectively, we computed an \textit{information
  score} for each voxel, reflecting the average reconstruction
accuracy across any patients with electrodes near each voxel-- where
the averages were weighted to reflect patients who had more electrodes
implanted near that location. This yielded a single map for each
dataset, highlighting regions that are potentially promising
implantation targets in terms of providing full-brain activity
information via SuperEEG (Fig.~\ref{fig:informap}A, B).  Despite task
and patient differences across the two datasets, we nonetheless found
that the maps of the most promising implantation targets derived from
both datasets were similar (voxelwise correlation between information
scores across the two datasets: $r = 0.18, p < 10^{-10}$).  While the
correspondence between the two maps was imperfect, our finding that
there were some commonalities between the two maps lends support to
the notion that different brain areas are differently informative
about full-brain activity patterns.  We also examined the intersection
between the top 10\% most informative voxels across the two datasets
(gray areas in Fig.~\ref{fig:informap}C).
Supporting the notion that structures that are highly interconnected
with the rest of the brain might be especially good targets for
implantation, this intersecting set of voxels with the highest
information scores included major portions of the dorsal attention
network (e.g., inferior parietal lobule, precuneus, inferior temporal
gyrus, thalamus, and striatum) as well as some portions of the default
mode network (e.g., angular gyrus) that are highly interconnected with
a large proportion of the brain's gray matter~\citep[e.g., ][]{TomaVolk11}.

@@

We also wondered what frequency properties contributed to the
reconstruction accuracy. To recover frequency-band specific voltage activity
from each recording, a second-order Butterworth filter was applied
across bands $\delta$, $\theta$, $\alpha$, $\beta$, $\gamma_L$, and
$\gamma_H$ with respective frequency ranges 2-4 Hz, 4-8 Hz, 8-12 Hz,
12-30 Hz, 30-60 Hz, and 60-100 Hz. Bandpass signals contain
information about the phase and amplitude of the underlying frequency
band, providing the most detailed view of the activity at that
frequency. Additionally, we calculated the reconstruction accuracies
using power at each band by taking the absolute
value of the Hilbert tranform of the bandpassed signal (see Fig.~\freqpower). To compute the broadband power (BB), we
first computed instantaneous power at 50 log-spaced frequencies
between 2 and 100 Hz using Morlet wavelets (n=4). Then, for each time
point and electrode, we fitted a robust Huber regression to frequency
and power in log-space, and calculated the mean height of the
resulting line. We then followed the same procedure outlined above
(see Fig.~\ref{fig:methods}) to
obtain the reconstruction accuracy at each bandpassed frequency as well
as broadband power for both datasets (see Fig.~\ref{fig:frequency}).
We found
the best average reconstruction accuracy for each dataset was obtained using broadband
power, which is a correlate of neural firing
~\citep{MannEtal09}. We also found a similar trend in reconstruction
accuracy for frequency bands across the two datasets, performing better at lower
frequencies (peaking at $\alpha$) and worst at high frequencies.

@@

We also wanted to see how the most informative regions changed by
frequency. Lower frequency oscillations, such as delta, theta, and alpha, have
been posited to be processing modes which span large swaths of
cortical regions and are thought to serve the purpose of integration
across many cortical sites ~\citep{Knya12}. Higher frequency
oscillations such as beta and gamma, are more limited to topographic
area ~\citep{Nune95}, and are modulated by slow oscillations
hierarchically ~\citep{EngeEtal10}. Using the frequency-specific reconstruction accuracies, we
followed the process outlined in Fig.~\ref{fig:informap}C, to obtain
the most informative locations for each frequency.  When we examined the intersection
between the top 10\% most informative voxels across the two datasets
for each bandpassed frequency, we found a somewhat consistent hierarchical
structure (see Fig.~\ref{fig:networks}E).

We then wondered which
networks were involved in each of
these frequency-specific intersecting regions.  To better understand the networks involved
we colored the regions (Fig.~\ref{fig:networks}A) based on the
network parcellation ~\citep[using $k$=7;][]{YeoEtal11}} (Fig.~\ref{fig:networks}D) and calculated the proportion of each
of those networks (Fig.~\ref{fig:networks}B).  Consistent with the idea outlined
above, we found networks most involved in the lowest frequency band,
$\delta$, were networks involved integration and global processing
(specifically the dorsal attention ~\citep{MarzEtal13}, default mode,
and frontoparietal networks).  Although these networks are also seen
in the $\theta$ intersection, they are less proportionally to the
visual network.  Alpha power is consistently modulated in EEG studies
of attention and visuo-motor ~\citep{MartiEtal10, MarzEtal13}, and we
similarly found a high proportion of visual and somatomotor networks
within this band.  We found $\beta$ intersections to predominantly
consist of somatomotor network, consistent with~\citep{RittEtal08,
  VanEEtal11}.  Also consistent with heavily cited
literature~\citep{FrieEtal07}, the network most predominant in the
$\gamma$ intersection was the visual network.  Lastly, broadband power
was found predominantly in frontoparietal and default mode networks,
consistent with ~\{OssaEtal11}.  We then carried out a
meta analysis using Neurosynth~\citep{RubiEtal17} to identify the 5
terms most commonly associated with the given map
(Fig.~\ref{fig:networks}C).

@@

\section*{Author Contributions}
J.R.M conceived and initiated the project. L.L.W.O. and
A.C.H. performed the analyses. J.R.M. and L.L.W.O. wrote the
manuscript.

>>>>

We first applied fourth order Butterworth notch filter to remove 60 Hz
($\pm$ .5 Hz) line noise from every recording (from every electrode).
Next, we downsampled the recordings (regardless of the original
samplerate) to 250 Hz.  (This downsampling step served to both
normalize for differences in sampling rates across patients and to
ease the computational burden of our subsequent analyses.)  We then
excluded any electrodes that showed putative epileptiform activity.
Specifically, we excluded from further analysis any electrode that
exhibited an maximum kurtosis of 10 or greater across all of that
patient's recording sessions.  We also excluded any patients with
fewer than 2 electrodes that passed this criteria, as the SuperEEG
algorithm requires measuring correlations between 2 or more electrodes
from each patient.  For Dataset 1, this yielded clean recordings from
4168 electrodes implanted throughout the brains of 67 patients
(Fig.~\ref{fig:methods}A); for Dataset 2, this yielded clean
recordings from 5023 electrodes from 78 patients.  Each individual
patient contributes electrodes from a limited set of brain locations,
which we localized in a common space~\citep[MNI152;][]{GrabEtal06}; an
example Dataset 1 patient's 54 electrodes that passed the above
kurtosis threshold test are highlighted in black and red.

@@

A second mechanism whereby a given region $x$ can contribute to the
recording at $\eta$ is through (direct or indirect) anatomical
connections between structures near $x$ and $\eta$.  Although
anatomical and functional correlations can differ
markedly~\citep[e.g.,][]{AdacEtal12, HoneEtal09, GoniEtal14}, we use temporal
correlations in the data to estimate these anatomical
connections~\citep{BeckEtal18}.  Let $\bar{R}$ be the set of locations
at which we wish to estimate local field potentials, and let
$R_{s} \subseteq \bar{R}$ be set of locations at which we observe local
field potentials from patient $s$ (excluding the electrodes that did
not pass the kurtosis test described above). In the analyses below we
define $\bar{R} = \cup_{s=1}^S R_s$.  We can calculate the expected
inter-electrode correlation matrix for patient $s$, where
$C_{s,k}(i,j)$ is the correlation between the time series of voltages
for electrodes $i$ and $j$ from subject $s$ during session $k$, using:

@@

We designed our approach to be agnostic to electrode impedances, as
electrodes that do not exist do not have impedances.  Therefore our
algorithm recovers voltages in standard deviation ($z$-scored) units
rather than attempting to recover absolute voltages. (This property
reflects the fact that $\hat{K}_{\beta_s, \alpha_s}$ and
$\hat{K}_{\alpha_s, \alpha_s}$ are correlation matrices rather than
covariance matrices.)  Also, we note that
Equation~\ref{eqn:reconstruction} requires computing a $T$ by $T$
matrix, which can become computationally intractable when $T$ is very
large (e.g., an example patient in Dataset 1, $T =
12786750$). However, because Equation~\ref{eqn:reconstruction} is time
invariant, we may compute $Y_{s,k,\beta_s}$ in a piecewise
manner by filling in $Y_{s,k,\beta_s}$ one row at a time (using the
corresponding samples from $Y_{s, k, \alpha_s}$).

@@

Although both datasets we examined provide good full-brain coverage
(when considering data from every patient; e.g.\
Fig.~\ref{fig:corrmap}C, D), electrodes are not placed uniformly
throughout the brain.  For example, electrodes are more likely to be
implanted in regions like the medial temporal lobe (MTL), and are
rarely implanted in occipital cortex (Fig.~\ref{fig:density}A, B).
Separately for each dataset, for each voxel in the 4~mm$^3$ voxel
MNI152 brain, we computed the proportion of electrodes in the dataset
that were contained within a 20 MNI unit radius sphere centered on
that voxel.  We defined the \textit{density} at that location as this
proportion.  Across Datasets 1 and 2, the electrode placement
densities were similar (correlation by voxel:
$r = 0.6, p < 10^{-10}$).  We wondered whether regions with good
covererage might be associated with better reconstruction accuracy
(e.g. Fig.~\ref{fig:corrmap}C and D indicate that many electrodes in the
MTL have relatively high reconstruction accuracy, and occipital
electrodes tend to have relatively low reconstruction accuracy).  To
test whether this held more generally across the entire brain, for
each dataset we computed the electrode placement density for each
electrode from each patient (using the proportion of \textit{other}
patients' electrodes within 20 MNI units of the given electrode).  We
then correlated these density values with the across-patient
reconstruction accuracies for each electrode.  We found a negative
correlation between reconstruction accuracy and density for Dataset 1,
but was unreliable for Dataset 2 (Dataset 1: $r = -0.21, p = 0.05$; Dataset 2:
$r = -0.30, p = 0.15$).  This indicates that the reconstruction
accuracies we observed are not driven by sampling density, but
rather may also reflect higher order properties of neural dynamics
such as functional correlations between distant
voxels~\citep{BetzEtal17b}.

@@

To gain insights into whether particular electrode locations might be
especially informative, we first computed the average reconstruction
accuracy across all of each patient's electrodes (using the
across-patients cross validation test; black histograms in
Fig.~\ref{fig:corrmap}A and B).  We labeled each patient's electrodes
in each dataset with the average reconstruction accuracy for that
patient.  In other words, we assigned every electrode from each given
patient the same value, reflecting how well the activity patterns at
those electrodes were reconstructed on average.  Next, for each voxel
in the 4~mm$^3$ MNI brain, we computed the average value across any
electrode (from any patient) that came within 20 MNI units of that
voxel's center.  Effectively, we computed an \textit{information
  score} for each voxel, reflecting the average reconstruction
accuracy across any patients with electrodes near each voxel-- where
the averages were weighted to reflect patients who had more electrodes
implanted near that location. This yielded a single map for each
dataset, highlighting regions that are potentially promising
implantation targets in terms of providing full-brain activity
information via SuperEEG (Fig.~\ref{fig:informap}A and B).  Despite task
and patient differences across the two datasets, we nonetheless found
that the maps of the most promising implantation targets derived from
both datasets were similar (voxelwise correlation between information
scores across the two datasets: $r = 0.18, p < 10^{-10}$).  While the
correspondence between the two maps was imperfect, our finding that
there were some commonalities between the two maps lends support to
the notion that different brain areas are differently informative
about full-brain activity patterns.  We also examined the intersection
between the top 10\% most informative voxels across the two datasets
(gray areas in Fig.~\ref{fig:informap}C, networks shown in
Fig.~\ref{fig:networks}A (All)).
Supporting the notion that structures that are highly interconnected
with the rest of the brain might be especially good targets for
implantation, this intersecting set of voxels with the highest
information scores included major portions of the dorsal attention
network (e.g., inferior parietal lobule, precuneus, inferior temporal
gyrus, thalamus, and striatum) as well as some portions of the default
mode network (e.g., angular gyrus) that are highly interconnected with
a large proportion of the brain's gray matter~\citep[e.g., ][]{TomaVolk11}.

@@

We also wondered what frequency properties contributed to the
reconstruction accuracy. To recover frequency-band specific voltage activity
from each recording, a second-order Butterworth filter was applied
across bands $\delta$, $\theta$, $\alpha$, $\beta$, $\gamma_L$, and
$\gamma_H$ with respective frequency ranges 2-4 Hz, 4-8 Hz, 8-12 Hz,
12-30 Hz, 30-60 Hz, and 60-100 Hz. Bandpass signals contain
information about the phase and amplitude of the underlying frequency
band, providing the most detailed view of the activity at that
frequency. Additionally, we calculated the reconstruction accuracies
using power at each band by taking the absolute
value of the Hilbert tranform of the bandpassed signal (see Fig.~\freqpower). To compute the broadband power (BB), we
first computed instantaneous power at 50 log-spaced frequencies
between 2 and 100 Hz using Morlet wavelets (n=4). Then, for each time
point and electrode, we fitted a robust Huber regression to frequency
and power in log-space, and calculated the mean height of the
resulting line. We then followed the same procedure outlined above
(see Fig.~\ref{fig:methods}) to
obtain the reconstruction accuracy at each bandpassed frequency as well
as broadband power for both datasets (see Fig.~\ref{fig:frequency}).
We found
the best average reconstruction accuracy for each dataset was obtained using broadband
power, which is a correlate of neural firing
~\citep{MannEtal09}. We also found a similar trend in reconstruction
accuracy for frequency bands across the two datasets, performing better at lower
frequencies (peaking at $\alpha$) and worst at high frequencies (see Fig.~\ref{fig:frequency}).

@@

We also wanted to see how the most informative regions changed by
frequency. Lower frequency oscillations, such as $\delta$, $\theta$, and $\alpha$, have
been posited to be involved in processing modes which span large swaths of
cortical regions and are thought to serve the purpose of integration
across many cortical sites ~\citep{Knya12}. Higher frequency
oscillations such as $\beta$ and $\gamma$, are more limited to topographic
areas ~\citep{Nune95}, and are modulated by slow oscillations
hierarchically ~\citep{EngeEtal10}. Using the frequency-specific reconstruction accuracies, we
followed the process outlined in Fig.~\ref{fig:informap}C, to obtain
the most informative locations for each frequency.  When we examined the intersection
between the top 10\% most informative voxels across the two datasets
for each bandpassed frequency, we found a somewhat consistent hierarchical
structure (see Fig.~\ref{fig:networks}E).

We then wondered which
networks were involved in each of
these frequency-specific intersecting regions.  To better understand the networks involved
we colored the regions (Fig.~\ref{fig:networks}A) based on the
network parcellation ~\citep[using $k$=7;][]{YeoEtal11}} (Fig.~\ref{fig:networks}D) and calculated the proportion of each
of those networks (Fig.~\ref{fig:networks}B).  Consistent with the idea outlined
above, we found networks most involved in the lowest frequency band,
$\delta$, were networks involved integration and global processing
(specifically the dorsal attention ~\citep{MarzEtal13}, default mode,
and frontoparietal networks).  Although these networks are also seen
in the $\theta$ intersection, they are less proportionally to the
visual network.  Alpha power is consistently modulated in EEG studies
of attention and visuo-motor ~\citep{MantEtal07, MarzEtal13}, and we
similarly found a high proportion of visual, somatomotor, and dorsal attentio networks
within this band.  We found $\beta$ intersections to predominantly
consist of somatomotor network, in line with~\citep{RittEtal09, VaneEtal11}.  Also accordant with heavily cited
literature~\citep{FrieEtal07}, the network most predominant in the
$\gamma$ intersection was the visual network.  Lastly, broadband power
was found predominantly in frontoparietal and default mode networks,
consistent with ~\citep{OssaEtal11}.

@@

\section*{Author Contributions}
J.R.M conceived and initiated the project. L.L.W.O., T.A.M, and
A.C.H. performed the analyses. J.R.M. and L.L.W.O. wrote the
manuscript.
